---
title: "tidyr_setup_and_play"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(readr)
library(tidyr)
```

```{r}
read_csv("/data/gapminder_data.csv") ->
  gapminder
gapminder
```


# From wide to long format with pivot_longer()
We can select the columns that will be pivoted by name . . .

```{r}
read_csv("/data/gapminder_wide.csv") ->
  gap_wide
gap_wide
```

```{r}
colnames(gap_wide)
```




```{r}
gap_wide %>%
  pivot_longer(
    cols= c(starts_with("pop"),
            starts_with("gdpPercap"),
            starts_with("lifeExp")),
    names_to = "obstype_year",
    values_to = "obs_values")
```



m_32
f_73
m_25

```{r}
gap_wide %>%
  pivot_longer(cols = c(-continent, -country), 
               names_to = "obstype_year",
               values_to = "obs_values") %>%
  separate(obstype_year, into = c("obs_type", "year"), sep="_") %>%
  mutate(year=as.integer(year)) ->
  gap_long
gap_long
```

# From long to intermediate
```{r}
gap_long %>%
  pivot_wider(names_from = obs_type, values_from = obs_values) ->
  gap_normal
gap_normal
```

```{r}
dim(gap_normal)
dim(gapminder)
```
```{r}
gapminder
```

```{r}
gap_normal %>%
  select(names(gapminder)) %>%
  arrange(country) %>%
  mutate(year=as.double(year))->
  gap_normal
gap_normal
```


```{r}
all.equal(gap_normal, gapminder)
```


```{r}
gap_wide %>%
  pivot_longer(
    cols = c(-continent, -country),
    names_to = "obstype_year", values_to = "obs_values"
  ) -> 
  gap_long
gap_long
```

We need to deal with `obstype_year`


```{r}
gap_long %>%
  separate(obstype_year, into = c("obs_type", "year"), sep="_") ->
  gap_long
gap_long$year = as.integer(gap_long$year)
gap_long
```

But this is problematic for two reasons:

1. Try running the above chunk a second time.  What happens? Why?
2. Look at the part where we make year into an integer. What is wrong with that

I prefer to do it all in one go (if possible), building it up line by line . . .

```{r}
gap_wide %>%
  pivot_longer(
    cols = c(-continent, -country),
    names_to = "obstype_year", values_to = "obs_values") %>%
  separate(obstype_year, into=c("obs_type", "year"), sep="_") %>%
  mutate(year = as.integer(year)) ->
  gap_long
```

# From long to intermediate format with pivot_wider()


