---
title: "tidyr_setup_and_play"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r eval=FALSE, include=FALSE}
library(readr)
download.file("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_wide.csv",
              "/data/gapminder_wide.csv")
```

```{r}
library(tidyr)
library(dplyr)
```


```{r}
gap_wide <- read.csv("/data/gapminder_wide.csv", stringsAsFactors = FALSE)
gap_wide
```

# From wide to long format with pivot_longer()
We can select the columns that will be pivoted by name . . .
```
gap_long = gap_wide %>%
  pivot_longer(
    cols = c(starts_with('pop'), 
             starts_with('lifeExp'), 
             starts_with('gdpPercap')),
    names_to = "obstype_year", 
    values_to = "obs_values")
gap_long
```
```{r}

```




OR sometimes it is easier to say which ones won't be pivoted
```
gap_wide %>%
  pivot_longer(
    cols = c(-continent, -country),
    names_to = "obstype_year", values_to = "obs_values"
  ) -> 
  gap_long
gap_long
```

We need to deal with `obstype_year`

```{r}

```

```
gap_long %>%
  separate(obstype_year, into = c("obs_type", "year"), sep="_") ->
  gap_long
gap_long$year = as.integer(gap_long$year)
gap_long
```

```{r}

```

But this is problematic for two reasons:

1. Try running the above chunk a second time.  What happens? Why?
2. Look at the part where we make year into an integer. What is wrong with that

I prefer to do it all in one go (if possible), building it up line by line . . .


```{r}
gap_wide %>%
  pivot_longer(
    cols = c(-continent, -country),
    names_to = "obstype_year", values_to = "obs_values") %>%
  separate(obstype_year, into=c("obs_type", "year"), sep="_") %>%
  mutate(year = as.integer(year)) ->
  gap_long
```

# From long to intermediate format with pivot_wider()
```{r}
gap_long %>%
  pivot_wider(names_from = obs_type,
              values_from = obs_values) ->
  gap_normal
gap_normal
```

```{r}
gapminder = read.csv("/data/gapminder_data.csv", stringsAsFactors = FALSE)
```


```{r}
dim(gap_normal)
dim(gapminder)
```

```{r}
names(gapminder)
names(gap_normal)
```

```{r}
gap_normal %>%
  select(names(gapminder)) ->
  gap_normal
names(gap_normal)
```

```{r}
all.equal(gap_normal, gapminder)
```

```{r}
head(gapminder)
```

```{r}
head(gap_normal)
```


```{r}
gap_normal %>%
  arrange(country, year) ->
  gap_normal
all.equal(gap_normal, gapminder)
```

```{r}
gap_long %>%
  unite(ID_var, continent, country, sep="_") %>%
  unite(var_names, obs_type, year, sep="_") %>%
  pivot_wider(names_from = var_names, values_from = obs_values) %>%
  separate(ID_var, c("continent", "country"), sep="_") ->
  gap_wide_betterID

```

```{r}
all.equal(gap_wide, as.data.frame(gap_wide_betterID))
# str(gap_wide)
# str(gap_wide_betterID)
```

